Equalizer\\Programming and User Guide

### 2010-08-23 Developer/ProgrammingGuide/paper.tex
Update Programming Guide
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications based on
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.7.5 for Equalizer 0.9.1+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
    \hline
  \end{tabularx}}
\clearpage

@@

\begin{description}
\item[eq::base] The eq::base namespace provides C++ classes to abstract
  the underlying operating system and implements common helper
  functionality. One example is the \textsf{eq::base::Clock} class
  providing a high-resolution timer. Classes in this namespace are fully
  documented in the API documentation on the Equalizer website, and are
  not subject of this Programming Guide.
\item[eq::net] The Equalizer network layer provides basic functionality
  for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{Session} and \textsf{Object}. Please refer to
  \sref{sNetwork} for an introduction into the network layer, and to
  \sref{sNetObject} for distributed objects.
\item[eq::util] The eq::util namespace provides common utility classes, which
  often the usage of OpenGL functions. Most classes in this namespace are used
  by the the Equalizer client library, but are usable independently from
  Equalizer for application development.
\item[eq] The core namespace for the Equalizer client library. The
  classes and their relationship in this namespace closely model the
  configuration file format. The classes in the \textsf{eq} namespace
  are the main subject of this Programming Guide.
\end{description}

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/namespaces}
  {\caption{\label{fNamespaces}Namespace Layering for eqPly and eqServer}}
  \vspace{-3ex}
\end{wrapfigure}
Examples shipped with Equalizer are implemented in their own namespaces,
e.g., \textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing
classes from the \textsf{eq} namespace, with the occasional usage of
functionality from the \textsf{eq::net} and \textsf{eq::base} namespace.

The functionality of the Equalizer server is implemented in the
\textsf{eq::server} namespace. This namespace does not yet expose an API
intended to be used by developers.

@@

Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchronization or protection of shared data.

@@

All classes in the example are in the \textsf{eqPly} namespace to avoid
type name ambiguities, in particular for the \textsf{Window} class which
is frequently used as a type in the global namespace by windowing
systems. \fig{fUml} shows how the most important Equalizer classes are
used through inheritance by the \textsf{eqPly} example.

The \textsf{eqPly} classes fall into two categories: Subclasses of the
rendering entities introduced in \sref{sConfig}, and classes for
distributing data. The function and typical usage for each of the
rendering entities is discussed in this section.

The distributed data classes are helper classes based on
\textsf{eq::Object} or its parent class \textsf{eq::net::Object}. They
illustrate the typical usage of distributed objects for static as well
as dynamic, frame-specific data. Furthermore they are used for a basic
scene graph distribution of the model data.

@@

Equalizer provides distributed objects which facilitate the
implementation of data distribution in a cluster
environment. Distributed objects are created by subclassing from
\textsf{eq::Object} or \textsf{eq::net::Object}. The application
programmer implements serialization and deserialization of the
distributed data. \sref{sNetObject2} covers distributed objects in detail.

@@

\begin{itemize}
\item Only the master instance of the object generates new versions, that is,
  \textsf{eq::net::Object::com\-mit} should only be called only on the master
  instance. It is possible to commit on slave instances, but special care has to
  be taken to handle possible conflicts. \sref{sSlaveCommit} covers slave object
  commits in detail.
\item Slave instance versions can only be advanced, that is,
  \textsf{eq::net::Object::sync( version )} with a version smaller than
  the current version will fail.
\item Newly mapped slave instances are mapped to the oldest available
  version by default, or to the version specified when calling
  \textsf{Session::mapObject}.
\end{itemize}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
  {\caption{\label{fUMLObject}eq::Object and eq::net::Object}}
\end{wrapfigure}
The creation of distributed, versioned objects is further facilitated by
\textsf{eq::Object}, which implements the most common way of tracking
data changes in versioned objects. The concept of a dirty bit mask is
used to mark parts of the object for serialization, while preserving the
capability to inherit objects. Other ways of implementing change
tracking, e.g., using incarnation counters, can still be implemented by
using \textsf{eq::net::Object} which leaves this flexibility to the
developer. \fig{fUMLObject} shows the relationship between
\textsf{eq::Object} and \textsf{eq::net::Object}.

The \textsf{FrameData} is sub-classed from \textsf{eq::Object}, and
consequently tracks its changes by setting the appropriate dirty bit
whenever it is changed. The serialization methods are called by the
\textsf{eq::Object} with the dirty bit mask needed to serialize all
data, or with the dirty bit mask of the changes since the last
\textsf{commit}. The \textsf{FrameData} only defines its own dirty bits
and serialization code:

{\footnotesize\begin{lstlisting}
/** The changed parts of the data since the last pack(). */
enum DirtyBits
{
    DIRTY_CAMERA  = eq::Object::DIRTY_CUSTOM << 0,
    DIRTY_FLAGS   = eq::Object::DIRTY_CUSTOM << 1,
    DIRTY_VIEW    = eq::Object::DIRTY_CUSTOM << 2,
    DIRTY_MESSAGE = eq::Object::DIRTY_CUSTOM << 3,
};

void FrameData::serialize( eq::net::DataOStream& os, const uint64_t dirtyBits )
{
    eq::Object::serialize( os, dirtyBits );
    if( dirtyBits & DIRTY_CAMERA )
        os << _translation << _rotation << _modelRotation;
    if( dirtyBits & DIRTY_FLAGS )
        os << _modelID << _renderMode << _colorMode << _ortho << _statistics
           << _help << _wireframe << _pilotMode << _idleMode;
    if( dirtyBits & DIRTY_VIEW )
        os << _currentViewID;
    if( dirtyBits & DIRTY_MESSAGE )
        os << _message;
}

void FrameData::deserialize( eq::net::DataIStream& is,
                             const uint64_t dirtyBits )
{
    eq::Object::deserialize( is, dirtyBits );
    if( dirtyBits & DIRTY_CAMERA )
        is >> _translation >> _rotation >> _modelRotation;
    if( dirtyBits & DIRTY_FLAGS )
        is >> _modelID >> _renderMode >> _colorMode >> _ortho >> _statistics
           >> _help >> _wireframe >> _pilotMode >> _idleMode;
    if( dirtyBits & DIRTY_VIEW )
        is >> _currentViewID;
    if( dirtyBits & DIRTY_MESSAGE )
        is >> _message;
}
\end{lstlisting}}%>>

@@

To dispatch events, Equalizer 'pumps' the native events. On WGL and GLX,
this happens on each thread with windows, whereas on AGL it happens on
the main thread and each pipe thread. By default, Equalizer pumps these
events automatically for the application in-between executing task
methods.

The method \textsf{Pipe::useMessagePump} can be overridden to return
\textsf{false} to not enable message dispatch for the respective thread.

The method \textsf{useMessagePump} is called by Equalizer during
application and pipe thread initialization (before configInit!) to
determine if the pipe's thread shall automatically dispatch OS
events. For AGL, this affects the node threads and pipe thread, since
the node thread message pump needs to dispatch the events to the pipe
thread event queue.

@@

Layout API entities (Canvas, Segment, Layout, View) are sub-classed like
all other Equalizer entities using the \textsf{NodeFactory}. Equalizer
registers the master instance of these entities on the application node,
so that the application can change mutable parameters and distribute its
own data to the render clients. \fig{fViewUML} shows the UML class
hierarchy for the \textsf{eqPly::View}.

The render clients can access a slave instance of the view using
\textsf{Channel::getView}. When called from one of the frame task
methods, this method will return the view of the current destination
channel for which the task method is executed. Otherwise it returns the
channel's native view, if it has one. Only destination channels of an
active canvas have a native view.

@@

The most common entity to subclass is the \textsf{View}, since the
application amends it by view-specific data. The view, like all other
layout entities, is indirectly sub-classed from \textsf{eq::Object}
through \textsf{eq::Frustum}. The design and usage of
\textsf{eq::Object} is explained in \sref{sEqObject}.

In \textsf{eqPly}, the application-specific data is the model
identifier. A new dirty bit for the model is defined, and the model
identifier is serialized when requested:

{\footnotesize\begin{lstlisting}
/** The changed parts of the view. */
enum DirtyBits
{
    DIRTY_MODEL       = eq::View::DIRTY_CUSTOM << 0,
};

void View::serialize( eq::net::DataOStream& os, const uint64_t dirtyBits )
{
    eq::View::serialize( os, dirtyBits );
    if( dirtyBits & DIRTY_MODEL )
        os << _modelID;
}

void View::deserialize( eq::net::DataIStream& is, const uint64_t dirtyBits )
{
    eq::View::deserialize( is, dirtyBits );
    if( dirtyBits & DIRTY_MODEL )
        is >> _modelID;
}

void View::setModelID( const uint32_t id )
{
    _modelID = id;
    setDirty( DIRTY_MODEL );
}
\end{lstlisting}}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/viewUML.pdf}
  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
\end{wrapfigure}
Changing view frustum parameters is typically done for non-fullscreen
application windows and multi-view layouts, where the rendering is not
meant to be viewed in real-world size in an immersive environment. A
typical use case is changing the field-of-view of the rendering.

@@

The \textsf{eq::net::Session} gathers provides higher-level
functionality to a set of nodes. The master instance of a session is
registered to a node, which makes this node the session server and
assigns a node-unique identifier to the session. All other nodes map
their instance of the session to their local listening node using the
session's identifier. One given node can participate in any number of
sessions, both as master and slave. This allows the flexible usage of
sessions to group different aspects of an application.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
  {\caption{\label{fNetSession}Communication between two Sessions}}
\end{wrapfigure}
Sessions instances can communicate to each other directly using an
\textsf{eq::net::Ses\-sion\-Packet}, which is derived from the
\textsf{NodePacket} and extends it by the session identifier. The
session ID is used by the node to dispatch the packet to the session.

@@

Distributed objects are created by subclassing from \textsf{eq::Object} or
\textsf{eq::net::Object}. The application programmer implements serialization
and deserialization of the distributed data. Objects are dynamically assigned to
a session, which assigns session-unique identifiers to objects and manages the
network communication between different instances of the same distributed
object.

@@

\begin{wrapfigure}{r}{.382\textwidth}
%\begin{figure}[h!t]
  \subfigure[]{\label{fObjectDist1}
    \includegraphics[width=.382\textwidth]{images/umlObjectSubclass}}\hfil
  \subfigure[]{\label{fObjectDist2}
    \includegraphics[width=.382\textwidth]{images/umlObjectProxy}}\hfil
  \subfigure[]{\label{fObjectDist3}
    \includegraphics[width=.382\textwidth]{images/umlObjectMI}}%
  {\caption{\label{fObjectDist}Object Distribution using Subclassing,
      Proxies or Multiple Inheritance}}
\end{wrapfigure}
There are multiple ways of implementing data distribution for an
existing C++ class hierarchy, such as a scene graph:
\begin{description}
\item[Subclassing] The classes to be distributed inherit from
  \textsf{eq::net::Object} and implement the serialization methods. This
  approach is recommended if the source code of existing classes can be
  modified. It is used for \textsf{eqPly::InitData} and
  \textsf{eqPly::FrameData}. (\fig{fObjectDist1})
\item[Proxies] For each object to be distributed, a proxy object is
  created which manages data distribution for its associated
  object. This requires the application to track changes on the object
  separately from the object itself. The model data distribution of
  eqPly is using this pattern. (\fig{fObjectDist2})
\item[Multiple Inheritance] A new class inheriting from the class to be
  distributed and from \textsf{eq::net::Object} implements the data
  distribution. This requires the application to instantiate a different
  type of object instead of the existing object, and to create wrapper
  methods in the superclass calling the original method and setting the
  appropriate dirty flags. This pattern is not used in
  eqPly. (\fig{fObjectDist3})
\end{description}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
\end{wrapfigure}
Versioned slave objects may also commit data to the master instance. However,
this requires the implementation to take special care to resolve conflicts
during deserialization. The algorithm and constraints for slave object commits
are explained in this section.

When using \textsf{commit} on slave object instances, multiple modifications on
the same database may happen simultaneously. In order to avoid excessive
synchronization in the network layer, Equalizer a light-weight approach which
requires the application to avoid or handle conflicting changes on the same
object.

@@

\subsubsection{\label{sEqObject}eq::Object}

Equalizer uses the \textsf{eq::net::Object} data distribution mechanism
for implementing data distribution between the server, application and
render clients. To do this, \textsf{eq::Object} implements one usage
pattern for the \textsf{eq::net::Object}, allowing inheritance with data
distribution. It is easier to use, but imposes one typical way to implement data
distribution.

The \textsf{eq::Object} data distribution is based on the concept of
dirty bits. Dirty bits form a 64-bit mask which marks which parts of the
object have to be distributed during the next commit.

For serialization, the default \textsf{eq::net::Object} serialization
functions are implemented by \textsf{eq::Object}, which (de-)serializes
and resets the dirty bits, and calls \textsf{serialize} or
\textsf{deserialize} with the bit mask specifying which data has to be
transmitted or received. During a commit or sync, the current dirty bits
are given, whereas during object mapping all dirty bits are passed to
the serialization methods.

To use \textsf{eq::Object}, the following steps have to be taken:

\begin{description}
\item[Inherit from \textsf{eq::Object}:] The base class will provide the
  dirty bit management and call serialize and deserialize
  appropriately. By overriding \textsf{getChangeType}, the default
  versioning strategy might be changed.
\item[Define new dirty bits:] Define dirty bits for each data item by
  starting at \textsf{Object::DIRTY\_CUSTOM}, shifting this value
  consecutively by each new dirty bit.
\item[Implement serialize and deserialize:] For each object-specific
  dirty bit which is set, stream the corresponding data item to or from
  the provided stream. Call the parent method first in both
  functions. For application-specific objects, write a (de-)serialization
  function.
\item[Mark dirty data:] In each 'setter' method, call \textsf{setDirty}
  with the corresponding dirty bit.
\end{description}

The inheritance chain \textsf{eq::View} $\rightarrow$
\textsf{eq::Frustum} $\rightarrow$ \textsf{eq::Object} $\rightarrow$
\textsf{eq::net::Object} serves as sample implementation for this
process. The \textsf{eq::Object} base class provides the facility to set
a name on the object, which is automatically distributed.

>>>>

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications based on
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.7.6 for Equalizer 0.9.1+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.7.0   & Sep 09, 2009 & Updates based on hardcopy print\\
    1.7.1   & Oct 20, 2009 & Updates for new eq::GLWindow\\
    1.7.2   & Nov 18, 2009 & Add \sref{sSubpixel} and \sref{sFSAA} for Subpixel compounds\\
    1.7.3   & Feb 02, 2010 & Revision for Equalizer 0.9.1\\
    1.7.4   & Feb 22, 2010 & Add \sref{sSlaveCommit}\\
    1.7.5   & Jun 03, 2010 & Add \sref{sClusterSetup}\\
    1.7.6   & Aug 23, 2010 & General update for fabric changes\\
    \hline
  \end{tabularx}}
\clearpage

@@

\begin{description}
\item[eq] The core namespace for the Equalizer client library. The
  classes and their relationship in this namespace closely model the
  configuration file format. The classes in the \textsf{eq} namespace
  are the main subject of this Programming Guide.
\item[eq::util] The eq::util namespace provides common utility classes, which
  often simplify the usage of OpenGL functions. Most classes in this namespace
  are used by the the Equalizer client library, but are usable independently
  from Equalizer for application development.
\item[eq::fabric] The eq::fabric namespace is the common data management and
  transport layer between the client library, the server and the upcoming
  administrative API. Most Equalizer client classes inherit from base classes in
  this namespace and have all their data access methods in these base
  classes. Furthermore, this namespace contains a \textsf{Serializable} class
  which implements one common use pattern of object-based data distribution
  using \textsf{eq::net::Object}.
\item[eq::net] The Equalizer network layer provides basic functionality
  for network communication, such as \textsf{Connection} and
  \textsf{ConnectionSet}, as well as higher-level functionality such as
  \textsf{Node}, \textsf{Session} and \textsf{Object}. Please refer to
  \sref{sNetwork} for an introduction into the network layer, and to
  \sref{sNetObject} for distributed objects.
\item[eq::base] The eq::base namespace provides C++ classes to abstract
  the underlying operating system and implements common helper
  functionality. One example is the \textsf{eq::base::Clock} class
  providing a high-resolution timer. Classes in this namespace are fully
  documented in the API documentation on the Equalizer website, and are
  not subject of this Programming Guide.
\end{description}

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/namespaces}
  {\caption{\label{fNamespaces}Namespace Layering for eqPly and eqServer}}
\end{wrapfigure}
Examples shipped with Equalizer are implemented in their own namespaces, e.g.,
\textsf{eqPly} or \textsf{eVolve}. They rely mostly on subclassing classes from
the \textsf{eq} namespace, with the occasional usage of functionality from the
\textsf{eq::util}, \textsf{eq::fabric}, \textsf{eq::net} and \textsf{eq::base}
namespace.

The functionality of the Equalizer server is implemented in the
\textsf{eq::server} namespace. The \textsf{eq::admin} namespace enable run-time
administration of Equalizer servers. These namespace does not yet expose an API
intended to be used by developers. \fig{fClientUml} provides an overview map of
the most important classes in the Equalizer namespace, grouped by functionality.

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/clientUML}
  {\caption{\label{fClientUml}Equalizer client UML map}}
\end{figure}

@@

Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model follows the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchronization or protection of shared data.

@@

\fig{fUml} shows how the most important Equalizer classes are used through
inheritance by the \textsf{eqPly} example. All classes in the example are in the
\textsf{eqPly} namespace to avoid type name ambiguities, in particular for the
\textsf{Window} class which is frequently used as a type in the global namespace
by windowing systems.

The \textsf{eqPly} classes fall into two categories: Subclasses of the
rendering entities introduced in \sref{sConfig}, and classes for
distributing data.

The function and typical usage for each of the rendering entities is discussed
in this section. Each of these classes inherits from a base class in the
\textsf{fabric} name\-space, which implements data distribution for the
entity. The fabric base classes are omitted in \fig{fUml}.

The distributed data classes are helper classes based on
\textsf{eq::fabric::Serializable} or its base class
\textsf{eq::net::Object}. They illustrate the typical usage of distributed
objects for static as well as dynamic, frame-specific data. Furthermore they are
used for a basic scene graph distribution of the model data.

@@

Equalizer provides distributed objects which facilitate the implementation of
data distribution in a cluster environment. Distributed objects are created by
subclassing from \textsf{eq::fabric::Serializable} or
\textsf{eq::net::Object}. The application programmer implements serialization
and deserialization of the distributed data. \sref{sNetObject2} covers
distributed objects in detail.

@@

\begin{itemize}
\item Only the master instance of the object generates new versions, that is,
  \textsf{commit} should only be called only on the master instance. It is
  possible to commit on slave instances, but special care has to be taken to
  handle possible conflicts. \sref{sSlaveCommit} covers slave object commits in
  detail.
\item Slave instance versions can only be advanced, that is, \textsf{sync(
  version )} with a version smaller than the current version will fail.
\item Newly mapped slave instances are mapped to the oldest available
  version by default, or to the version specified when calling
  \textsf{Session::mapObject}.
\end{itemize}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/umlObject.pdf}
  {\caption{\label{fUMLObject}eq::fabric::Serializable and eq::net::Object}}
\end{wrapfigure}
The creation of distributed, versioned objects is simplified when using
\textsf{eq::fabric::Serializable}, which implements one common way of tracking
data changes in versioned objects. The concept of a dirty bit mask is used to
mark parts of the object for serialization, while preserving the capability to
inherit objects. Other ways of implementing change tracking, e.g., using
incarnation counters, can still be implemented by using \textsf{eq::net::Object}
which leaves all flexibility to the developer. \fig{fUMLObject} shows the
relationship between \textsf{eq::fabric::Serializable} and
\textsf{eq::net::Object}.

The \textsf{FrameData} is sub-classed from \textsf{Serializable}, and
consequently tracks its changes by setting the appropriate dirty bit
whenever it is changed. The serialization methods are called by the
\textsf{eq::fabric::Serializable} with the dirty bit mask needed to serialize all
data, or with the dirty bit mask of the changes since the last
\textsf{commit}. The \textsf{FrameData} only defines its own dirty bits
and serialization code:

{\footnotesize\begin{lstlisting}
/** The changed parts of the data since the last pack(). */
enum DirtyBits
{
    DIRTY_CAMERA  = eq::fabric::Serializable::DIRTY_CUSTOM << 0,
    DIRTY_FLAGS   = eq::fabric::Serializable::DIRTY_CUSTOM << 1,
    DIRTY_VIEW    = eq::fabric::Serializable::DIRTY_CUSTOM << 2,
    DIRTY_MESSAGE = eq::fabric::Serializable::DIRTY_CUSTOM << 3,
};

void FrameData::serialize( eq::net::DataOStream& os, const uint64_t dirtyBits )
{
    eq::fabric::Serializable::serialize( os, dirtyBits );
    if( dirtyBits & DIRTY_CAMERA )
        os << _translation << _rotation << _modelRotation;
    if( dirtyBits & DIRTY_FLAGS )
        os << _modelID << _renderMode << _colorMode << _ortho << _statistics
           << _help << _wireframe << _pilotMode << _idleMode;
    if( dirtyBits & DIRTY_VIEW )
        os << _currentViewID;
    if( dirtyBits & DIRTY_MESSAGE )
        os << _message;
}

void FrameData::deserialize( eq::net::DataIStream& is,
                             const uint64_t dirtyBits )
{
    eq::fabric::Serializable::deserialize( is, dirtyBits );
    if( dirtyBits & DIRTY_CAMERA )
        is >> _translation >> _rotation >> _modelRotation;
    if( dirtyBits & DIRTY_FLAGS )
        is >> _modelID >> _renderMode >> _colorMode >> _ortho >> _statistics
           >> _help >> _wireframe >> _pilotMode >> _idleMode;
    if( dirtyBits & DIRTY_VIEW )
        is >> _currentViewID;
    if( dirtyBits & DIRTY_MESSAGE )
        is >> _message;
}
\end{lstlisting}}%>>

@@

To dispatch events, Equalizer 'pumps' the native events. On WGL and GLX, this
happens on each thread with windows, whereas on AGL it happens on the main
thread and on each pipe thread. By default, Equalizer pumps these events
automatically for the application in-between executing task methods.

The method \textsf{Pipe::createMessagePump} is called by Equalizer during
application and pipe thread initialization (before \textsf{Pipe::configInit}!)
to create a message pump for the given pipe. For AGL, this affects the node
thread and pipe threads, since the node thread message pump needs to dispatch
the events to the pipe thread event queue. Custom message pumps may be
implemented, and it is valid to return no message pump to disable message
dispatch for the respective pipe.

@@

Layout API entities (Canvas, Segment, Layout, View) are sub-classed like all
other Equalizer entities using the \textsf{NodeFactory}. Equalizer registers the
master instance of these entities on the server node. Mutable parameters, e.g.,
the active layout of a canvas, are distributed using slave object commits
(cf. \sref{sSlaveCommit}). Application-specific data can be attached to a view
using a distributable \textsf{UserData} object. \fig{fViewUML} shows the UML
class hierarchy for the \textsf{eqPly::View}.

\begin{wrapfigure}{r}{.612\textwidth}
  \includegraphics[width=.612\textwidth]{images/viewUML.pdf}
  {\caption{\label{fViewUML}UML Hierarchy of \textsf{eqPly::View}}}
\end{wrapfigure}
The render clients can access a slave instance of the view using
\textsf{Channel::getView}. When called from one of the frame task
methods, this method will return the view of the current destination
channel for which the task method is executed. Otherwise it returns the
channel's native view, if it has one. Only destination channels of an
active canvas have a native view.

@@

The most common entity to subclass is the \textsf{View}, since the application
often amends with view-specific application data. A view might have a
distributable user data object, which has to inherit from
\textsf{eq::net::Object} to be committed and synchronized with the associated
object.

Externalizing the data distribution in a \textsf{UserData} object is necessary
since the server holds the master instance of a view. The server contains no
application code, and would not be able to (de-)serialize the application data
of a view. Instead, the user data object has its master and slave instances only
in application processes, and the server only handles the identifier and version
of this object.

In \textsf{eqPly}, the application-specific data is the model identifier and the
number of anti-aliasing steps to be rendered when idle. This data is distributed
by a \textsf{View::Proxy} object which serves as the view's user data. The proxy
defines the necessary dirty bits and serializes the data:

{\footnotesize\begin{lstlisting}
/** The changed parts of the view. */
enum DirtyBits
{
    DIRTY_MODEL       = eq::fabric::Serializable::DIRTY_CUSTOM << 0,
    DIRTY_IDLE        = eq::fabric::Serializable::DIRTY_CUSTOM << 1
};

void View::Proxy::serialize( eq::net::DataOStream& os,
                             const uint64_t dirtyBits )
{
    if( dirtyBits & DIRTY_MODEL )
        os << _view->_modelID;
    if( dirtyBits & DIRTY_IDLE )
        os << _view->_idleSteps;
}

void View::Proxy::deserialize( eq::net::DataIStream& is,
                               const uint64_t dirtyBits )
{
    if( dirtyBits & DIRTY_MODEL )
        is >> _view->_modelID;
    if( dirtyBits & DIRTY_IDLE )
    {
        is >> _view->_idleSteps;
        if( isMaster( ))
            setDirty( DIRTY_IDLE ); // redistribute slave settings
    }
}
\end{lstlisting}}

The \textsf{eqPly::View} sets its \textsf{Proxy} as the user data object in the
constructor. By default, the master instance of the view's user data is on the
application instance of the view. This may be changed by overriding
\textsf{hasMasterUserData}. The proxy object registration, mapping and
synchronization is fully handled be the \textsf{fabric} layer, no further
handling has to be done by the application:

{\footnotesize\begin{lstlisting}
View::View( eq::Layout* parent )
    : [...]
{
    setUserData( &_proxy );
}
void View::setModelID( const uint32_t id )
{
    _modelID = id;
    setDirty( DIRTY_MODEL );
}
\end{lstlisting}}

@@

Changing view frustum parameters is typically done for non-fullscreen
application windows and multi-view layouts, where the rendering is not
meant to be viewed in real-world size in an immersive environment. A
typical use case is changing the field-of-view of the rendering.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
  {\caption{\label{fNetSession}Communication between two Sessions}}
\end{wrapfigure}
The \textsf{eq::net::Session} gathers provides higher-level
functionality to a set of nodes. The master instance of a session is
registered to a node, which makes this node the session server and
assigns a node-unique identifier to the session. All other nodes map
their instance of the session to their local listening node using the
session's identifier. One given node can participate in any number of
sessions, both as master and slave. This allows the flexible usage of
sessions to group different aspects of an application.

@@

Sessions instances can communicate to each other directly using an
\textsf{eq::net::Ses\-sion\-Packet}, which is derived from the
\textsf{NodePacket} and extends it by the session identifier. The
session ID is used by the node to dispatch the packet to the session.

@@

Distributed objects subclass from \textsf{eq::fabric::Serializable} or
\textsf{eq::net::Object}. The application programmer implements serialization
and deserialization of the distributed data. Objects are dynamically assigned to
a session, which assigns session-unique identifiers to objects and manages the
network communication between different instances of the same distributed
object.

@@

%\begin{wrapfigure}{r}{.382\textwidth}
\begin{figure}[h!t]
  \subfigure[]{\label{fObjectDist1}
    \includegraphics[width=.3\textwidth]{images/umlObjectSubclass}}\hfil
  \subfigure[]{\label{fObjectDist2}
    \includegraphics[width=.3\textwidth]{images/umlObjectProxy}}\hfil
  \subfigure[]{\label{fObjectDist3}
    \includegraphics[width=.3\textwidth]{images/umlObjectMI}}%
  {\caption{\label{fObjectDist}Object Distribution using Subclassing,
      Proxies or Multiple Inheritance}}
\end{figure}
There are multiple ways of implementing data distribution for an
existing C++ class hierarchy, such as a scene graph:
\begin{description}
\item[Subclassing] The classes to be distributed inherit from
  \textsf{eq::net::Object} and implement the serialization methods. This
  approach is recommended if the source code of existing classes can be
  modified. It is used for \textsf{eqPly::InitData} and
  \textsf{eqPly::FrameData}. (\fig{fObjectDist1})
\item[Proxies] For each object to be distributed, a proxy object is
  created which manages data distribution for its associated
  object. This requires the application to track changes on the object
  separately from the object itself. The model data distribution of
  eqPly is using this pattern. (\fig{fObjectDist2})
\item[Multiple Inheritance] A new class inheriting from the class to be
  distributed and from \textsf{eq::net::Object} implements the data
  distribution. This requires the application to instantiate a different
  type of object instead of the existing object, and to create wrapper
  methods in the superclass calling the original method and setting the
  appropriate dirty flags. This pattern is not used in
  eqPly. (\fig{fObjectDist3})
\end{description}

@@

Versioned slave objects may also commit data to the master instance. However,
this requires the implementation to take special care to resolve conflicts
during deserialization. The algorithm and constraints for slave object commits
are explained in this section.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/slaveWrites.pdf}
  {\caption{\label{fSlaveCommit}Slave Commit Communication Sequence}}
\end{wrapfigure}
When using \textsf{commit} on slave object instances, multiple modifications on
the same database may happen simultaneously. In order to avoid excessive
synchronization in the network layer, Equalizer a light-weight approach which
requires the application to avoid or handle conflicting changes on the same
object.

@@

\subsubsection{\label{sEqObject}eq::fabric::Serializable}

Equalizer uses the \textsf{eq::net::Object} data distribution mechanism for
implementing data distribution between the server, application and render
clients. Based on dirty bits, \textsf{eq::fabric::Serializable} implements one
usage pattern for the \textsf{eq::net::Object}, allowing inheritance with data
distribution. It is easier to use, but imposes one typical way to implement data
distribution.

The \textsf{eq::fabric::Serializable} data distribution is based on the concept
of dirty bits. Dirty bits form a 64-bit mask which marks which parts of the
object have to be distributed during the next commit.

For serialization, the default \textsf{eq::net::Object} serialization functions
are implemented by \textsf{eq::fabric::Serializable}, which (de-)serializes and
resets the dirty bits, and calls \textsf{serialize} or \textsf{deserialize} with
the bit mask specifying which data has to be transmitted or received. During a
commit or sync, the current dirty bits are given, whereas during object mapping
all dirty bits are passed to the serialization methods.

To use \textsf{eq::fabric::Serializable}, the following steps have to be taken:

\begin{description}
\item[Inherit from \textsf{eq::fabric::Serializable}:] The base class will
  provide the dirty bit management and call serialize and deserialize
  appropriately. By overriding \textsf{get\-Chan\-ge\-Type}, the default
  versioning strategy might be changed.
\item[Define new dirty bits:] Define dirty bits for each data item by
  starting at \textsf{Serializable::DIRTY\_CUSTOM}, shifting this value
  consecutively by each new dirty bit.
\item[Implement serialize and deserialize:] For each object-specific
  dirty bit which is set, stream the corresponding data item to or from
  the provided stream. Call the parent method first in both
  functions. For application-specific objects, write a (de-)serialization
  function.
\item[Mark dirty data:] In each 'setter' method, call \textsf{setDirty}
  with the corresponding dirty bit.
\end{description}

The Equalizer entities use the common base class
\textsf{eq::fabric::Object}. This class manages common entity data, and is not
intended to by subclassed directly.  The inheritance chain \textsf{eq::View}
$\rightarrow$ \textsf{eq::fabric::View} $\rightarrow$ \textsf{eq::Object}
$\rightarrow$ \textsf{eq::fabric::Se\-rial\-izable} $\rightarrow$
\textsf{eq::net::Object} serves as sample implementation for this process.
